Impact of Pandemic Policies on Rural Indian Women's Mental Health and Economic Stability

### 2024-04-19 outputs/paper.qmd
variables

---
title: "Tracing Hidden Struggles of Rural Indian Women during Pandemic"
author: Ruoxian Wu
thanks: "Code and data are available at: https://github.com/ScarletWu/Tracing_Hidden_Struggles_of_Rural_Indian_Women_during_Pandemic.git. Replication on Social Science Reproduction platform is available at: https://www.socialsciencereproduction.org/reproductions/c37e126c-8a08-4a52-8f24-8c92a497fad5/index."
date: "April 16, 2024"
date-format: long
abstract: "This paper replicates a pivotal study on the socio-economic impacts of pandemic containment policies on women in India, specifically examining mental health and financial status. Utilizing a dataset spanning several indicators of economic and psychological health, this study employs a combination of linear and Lasso regression models to assess the implications of these measures. Key findings suggest significant correlations between containment intensity and various aspects of women's well-being, including mental health and economic stability. The results underscore the importance of considering gender-specific outcomes in policy-making processes, particularly in times of global crises. This study contributes to the broader discourse on public health interventions and their gendered impacts, advocating for more nuanced approaches in future public health strategies."
format: pdf
toc: true
number-sections: true
bibliography: references.bib
---

@@

The study "Women’s Well-being during a Pandemic and its Containment" from the Journal of Development Economics (2022) examines the effects of COVID-19 and related containment policies on women in lower-income countries @bau2022. My paper takes a close look at the original study's methodologies and outcomes. Using the data provided by the original authors, I replicated their research, applying fixed-effect analysis and regression models to assess how public health funding influences outcomes during health crises like COVID-19.

This analysis reviews the complex links between funding, policy effectiveness, and socio-economic impacts. It revisits the original study's data and approach, enhancing our understanding of how health initiatives can better serve vulnerable populations during emergencies.

The original paper highlighted the decline in mental health, financial status, and meal nutrition for women due to pandemic policies (@bau2022). My work corroborates these findings and examines the balance needed between public health interventions and socio-economic support. Echoing findings from the World Health Organization (@who2020) and studies like that of @xie2021, my analysis confirms that strategic health investments are pivotal for crisis mitigation and socio-economic stability.

This paper aims to inform policy that effectively utilizes health funding to improve outcomes and support those most affected by health crises, highlighting the need for policies that are responsive to the socio-economic realities of women in low-income countries during pandemics.

# Data
## Tools
My replication employed R [@r] along with several essential packages for data manipulation and visualization. Specifically, I utilized Haven [@haven] for importing and exporting data, Dplyr [@dplyr] for data manipulation, Ggplot2 [@ggplot2] for creating visualizations, Readr [@readr] for reading rectangular text data, Here [@here] for managing file paths, Janitor [@janitor] for examining and cleaning data, KableExtra [@kableExtra] for constructing complex tables, Knitr [@knitr] for dynamic report generation, and Tidyverse [@tidyverse] for a cohesive data science workflow. Additionally, I employed Coefplot [@coefplot] for visualizing regression coefficients and Broom [@broom], Lmtest [@lmtest], and Sandwich [@sandwich] for regression diagnostics and robust standard errors. For advanced modeling techniques, I utilized Glmnet [@glmnet] for Lasso regression.

## Source

This critical analysis utilizes replication data associated with the article "Women’s well-being during a pandemic and its containment" from the Journal of Development Economics. The data, along with associated code, needs to be requested from 'https://datasets.iza.org/dataset/1395/g2lm-lic-women-s-well-being-during-a-pandemic-and-its-containment' to facilitate the replication of key findings such as statistical models and graphical representations. By enabling the reproduction of the authors' analyses, this data contributes to the transparency and credibility of the study's conclusions. The replication package can be find and downloaded after requesting access.

@@

@@

```{r}
#| message: false
#| include: false
# Load data
# Load data
data_path <- "/cloud/project/data/raw_data/"
output_path <- "/cloud/project/outputs/"
data <- read_dta("/cloud/project/data/raw_data/covid_gender_data.dta")
```

@@

Figure 2 illustrates scatter plots that relate the containment measures to different outcomes, including anxiety, depression, exhaustion, lost income, reduced meals, and safety. The plots typically have the containment intensity on the x-axis and the outcome variable on the y-axis. The line represents the best-fit linear relationship between containment and the specific outcome.  In these analyses, a regression coefficient (β) is used to measure the strength and direction of the relationship, accompanied by standard errors (SE) to indicate the precision of these estimates. As containment measures become more stringent , individuals report increasing levels of depression (β = 0.295, @fig-2-1), exhaustion (β = 0.504,  @fig-2-2), and anxiety (β = 0.332,  @fig-2-3), alongside a decreased sense of safety (β = 0.282, @fig-2-4). Additionally, there is a notable reduction in meal frequency (β = 1.15, @fig-2-5) and an increase in the loss of individual income (β = 0.065, @fig-2-6) associated with stricter containment policies. These β values are all positive, indicating a direct correlation with containment levels, and the SE values (in parentheses) reflect the variability of the estimates within the clusters of districts.

The results depicted in the figure provide valuable insights into the relative importance of predictor variables across different models. By visually presenting the estimated coefficients and their confidence intervals, this plot offers a clear and direct interpretation of the model results. Unlike the original paper, which relied on tables to illustrate the model results, this coefficient figure provides a more intuitive and easy-to-understand representation. The prominence of certain predictor variables, as indicated by larger effect sizes and narrower confidence intervals, suggests their stronger associations with the outcomes of interest. These findings have important implications for understanding the drivers of the observed outcomes and can inform targeted interventions and policy decisions. Overall, the use of this coefficient figure enhances the accessibility and comprehensibility of the model results, facilitating informed decision-making and further research endeavors.

## Mental Health Deterioration
The study documents a significant deterioration in mental health among the surveyed women, with increases in reported depression, exhaustion, and anxiety. These findings align with broader literature indicating that women, particularly in lower-income settings, bear a disproportionate burden of mental health challenges during crises.

@@

To refine my models, I utilized logistic regression techniques, which  incorporated LASSO for variable selection. This approach was crucial in determining the strength and significance of relationships between the pandemic's containment measures and various mental health indicators. The results I uncovered were consistent with those of the original paper, indicating that containment measures such as lockdowns and restricted movements were associated with heightened feelings of depression and anxiety among women.

@@

## Economic and Nutritional Impact
Economically, the pandemic led to dramatic income losses and heightened food insecurity, with the latter particularly pronounced among women. The findings suggest that containment measures, while necessary for public health, inadvertently exacerbated these issues, underscoring the need for targeted support measures.

# Discussion
## Socioeconomic Strain and Mental Health
The data suggest that the socioeconomic strains induced by the pandemic, compounded by containment measures, have had a profound impact on women's mental health. This aligns with findings from other studies highlighting the disproportionate burden of mental health challenges on women during crises. For instance, (@smith2020) argue that women often bear the brunt of caregiving responsibilities and are more likely to experience stress and anxiety during crises due to increased caregiving demands and economic vulnerability. Similarly, (@nguyen2018) found that women, particularly those in low-income settings, are more susceptible to mental health issues during health emergencies, attributing this vulnerability to pre-existing socioeconomic disparities and limited access to resources. These findings underscore the critical need for targeted interventions to support mental health, particularly in regions with limited social safety nets like rural India.

@@

In conclusion, the analysis affirms the significant negative repercussions of the COVID-19 pandemic and associated containment measures on women's mental health and economic stability in rural India. It underscores the necessity of designing and implementing nuanced public health policies that account for the socioeconomic realities of vulnerable populations and prioritize their well-being.

>>>>

---
title: "Impact of Pandemic Policies on Rural Indian Women's Mental Health and Economic Stability"
author: Ruoxian Wu
thanks: "Code and data are available at: https://github.com/ScarletWu/Tracing_Hidden_Struggles_of_Rural_Indian_Women_during_Pandemic.git. Replication on Social Science Reproduction platform is available at: https://www.socialsciencereproduction.org/reproductions/c37e126c-8a08-4a52-8f24-8c92a497fad5/index."
date: "April 16, 2024"
date-format: long
abstract: "This study replicates a research on the effects of pandemic containment policies on women in India, focusing on their mental health and financial status. The study uses a dataset that covers various indicators of economic and psychological health. It employs a combination of linear and Lasso regression models to evaluate the implications of these measures. The results reveal significant correlations between containment intensity and various aspects of women's well-being, such as mental health and economic stability. The findings highlight the significance of considering gender-specific outcomes in policy-making processes, particularly during global crises. This study contributes to the broader conversation on public health interventions and their gendered impacts.
"
format: pdf
toc: true
number-sections: true
bibliography: references.bib
---

@@

This paper critically examines the impact of COVID-19 containment policies on rural Indian women and focuses on their mental health and financial stability. The original study titled "Women's Well-being during a Pandemic and its Containment" from the Journal of Development Economics (2022) laid a foundation by exploring the broader effects of the pandemic on women in lower-income countries. This paper replicates and extends the research to delve deeper into the specific challenges faced by women in rural India, a demographic that remains underrepresented in global discussions on public health and economic policy (@bau2022).

The necessity of this study stems from the unique socio-economic conditions of rural India, where women face distinct challenges that are exacerbated by crises like the COVID-19 pandemic. Previous research has highlighted how pandemics disproportionately affect women, particularly in terms of increased domestic burdens, reduced economic opportunities, and heightened mental health issues. By focusing on rural Indian women, this study contributes to a nuanced understanding of how containment measures, while necessary to curb the spread of the virus, have had unintended adverse effects on this vulnerable subgroup.

The methodology of this paper involves a detailed analysis of data originally collected by the study from the Journal of Development Economics, enhanced by additional variables and refined econometric models. This replication employs a combination of fixed-effect analysis and advanced regression techniques to provide new insights into the interplay between public health policies and women's well-being.

Through this structured analysis, the paper aims to inform effective policy-making that considers the socio-economic realities faced by rural Indian women. This study aims to ensure that future public health interventions are inclusive and equitable.

# Data
## Source and Provenance
This analysis utilizes replication data associated with the article "Women’s well-being during a pandemic and its containment" from the Journal of Development Economics. The data, along with associated code, needs to be requested from "https://datasets.iza.org/dataset/1395/g2lm-lic-women-s-well-being-during-a-pandemic-and-its-containment" to facilitate the replication of key findings such as statistical models and graphical representations.

## Data Collection and Cleaning
The data was collected through a collaborative effort between local non-governmental organizations and academic researchers across six states in rural India. This collaboration ensured that high-quality, reliable data was collected, which reflected the unique socio-economic conditions of the surveyed regions. The survey was conducted among households that were initially interviewed in the fall of 2019, providing a critical pre-pandemic baseline. This baseline enhances our analysis of the changes attributable to COVID-19 containment policies. Part of data for this study was collected via phone interviews in August 2020 to ensure safety during the pandemic. The interviews covered a wide range of topics including demographic information, economic factors, and health outcomes. The study focused on mental health indicators and nutrition levels, which were measured using validated psychological scales like PHQ9 and GAD7, and food security questions adapted from national health surveys ().

To ensure the robustness of the analysis, the initial phase of data cleaning involved several critical steps.  The overall data distribution is anaylzed to understand the dataset.Missing values were cleaned to prepare for the modeling. The cleaning process also involved standardizing variable formats across the dataset to ensure consistency in analysis.

## Tools
My replication employed R [@r] along with several essential packages for data manipulation and visualization. Specifically, I utilized Haven [@haven] for importing and exporting data, Dplyr [@dplyr] for data manipulation, Ggplot2 [@ggplot2] for creating visualizations, Readr [@readr] for reading rectangular text data, Here [@here] for managing file paths, Janitor [@janitor] for examining and cleaning data, KableExtra [@kableExtra] for constructing complex tables, Knitr [@knitr] for dynamic report generation, and Tidyverse [@tidyverse] for a cohesive data science workflow. Additionally, I employed Coefplot [@coefplot] for visualizing regression coefficients and Broom [@broom], Lmtest [@lmtest], and Sandwich [@sandwich] for regression diagnostics and robust standard errors. For advanced modeling techniques, I utilized Glmnet [@glmnet] for Lasso regression.

@@

### Model Variables:

- `$y_{iasd}$`: The outcome variable for an individual, such as mental health or economic status.
- `$\beta \text{containment}_d$`: Coefficient for containment policies in the individual's district, measuring the impact of these policies on the outcome.
- `$\alpha_a$`: Age fixed effects to control for age-related variations in the outcome.
- `$\delta_s$`: State fixed effects to account for differences attributable to the state of residence.
- `$\Gamma X_i$`: Effects of control variables, including indicators for the district being in a red or orange zone during specific lockdown periods.
- `$\epsilon_{iasd}$`: Error term capturing unobserved influences on the outcome.

@@

```{r}
#| message: false
#| include: false
# Load data
data_path <- "/cloud/project/data/raw_data/"
output_path <- "/cloud/project/outputs/"
data <- read_dta("/cloud/project/data/raw_data/covid_gender_data.dta")
```

@@

Figure 2 illustrates scatter plots that relate the containment measures to different outcomes, including anxiety, depression, exhaustion, lost income, reduced meals, and safety. The plots have the containment intensity on the x-axis and the outcome variable on the y-axis. The line represents the best-fit linear relationship between containment and the specific outcome.  In these analyses, a regression coefficient (β) is used to measure the strength and direction of the relationship, accompanied by standard errors (SE) to indicate the precision of these estimates. As containment measures become more stringent , individuals report increasing levels of depression (β = 0.295, @fig-2-1), exhaustion (β = 0.504,  @fig-2-2), and anxiety (β = 0.332,  @fig-2-3), alongside a decreased sense of safety (β = 0.282, @fig-2-4). Additionally, there is a notable reduction in meal frequency (β = 1.15, @fig-2-5) and an increase in the loss of individual income (β = 0.065, @fig-2-6) associated with stricter containment policies. These β values are all positive, indicating a direct correlation with containment levels, and the SE values (in parentheses) reflect the variability of the estimates within the clusters of districts.

@@

To refine my models, I utilized logistic regression techniques, which incorporated LASSO for variable selection. This approach was crucial in determining the strength and significance of relationships between the pandemic's containment measures and various mental health indicators. The results I uncovered were consistent with those of the original paper, indicating that containment measures such as lockdowns and restricted movements were associated with heightened feelings of depression and anxiety among women.The stricter containment measures correlate with poorer mental health and economic stability.

@@

The results depicted in the figure provide valuable insights into the relative importance of predictor variables across different models. By visually presenting the estimated coefficients and their confidence intervals, this plot offers a clear and direct interpretation of the model results. Unlike the original paper, which relied on tables to illustrate the model results, this coefficient figure provides a more intuitive and easy-to-understand representation.

# Discussion
## Mental Health Deterioration
The study documents a significant deterioration in mental health among the surveyed women, with increases in reported depression, exhaustion, and anxiety. These findings align with broader literature indicating that women, particularly in lower-income settings, bear a disproportionate burden of mental health challenges during crises. For instance, (@smith2020) argue that women often bear the brunt of caregiving responsibilities and are more likely to experience stress and anxiety during crises due to increased caregiving demands and economic vulnerability. Similarly, (@nguyen2018) found that women, particularly those in low-income settings, are more susceptible to mental health issues during health emergencies, attributing this vulnerability to pre-existing socioeconomic disparities and limited access to resources. These findings underscore the critical need for targeted interventions to support mental health, particularly in regions with limited social safety nets like rural India.

## Economic and Nutritional Impact
The study indicates that pandemics can cause economic instability, particularly affecting marginalized communities. While containment measures reduce the spread of COVID-19, they have inadvertently deepened economic disparities, leading to significant income reductions and compromised food security. This observation aligns with previous research, which has shown that economic shocks have a disproportionate impact on low-income populations, exacerbating poverty and reducing access to essential resources (@Patel2019Economic).

The economic strain has a cascading effect on nutrition, as evidenced by the increased instances of meal reductions among the affected populations. This link between economic instability and nutritional outcomes is well-documented, with studies illustrating how economic downturns directly impact dietary quality and overall health in rural settings (@Martin2020Economic). The nutritional deficits noted in this study underscore the need for integrated policy responses that address both the immediate and long-term economic and nutritional needs of vulnerable groups. Policies aimed at enhancing food security must be coupled with economic support measures to effectively counteract the adverse effects of pandemic-related disruptions. These interventions should prioritize the distribution of food aid, support for local agriculture, and programs that provide direct financial support to those most at risk of economic and nutritional instability.

@@

# Conclusion
In conclusion, the analysis affirms the significant negative repercussions of the COVID-19 pandemic and associated containment measures on women's mental health and economic stability in rural India. The findings emphasize the pressing need for public health policies that are not only effective in controlling the spread of diseases but also considerate of the socioeconomic realities of the populations they impact. By taking these factors into account while making policies, we can strive to not only manage public health crises more efficiently but also reduce their harmful impacts on vulnerable populations.
